---
title: "Habitat Suitability Models for Diadromous Fish in the Northeast United States"
author: "Vanessa Mahan"
date: "2023-06-15"
output:
  word_document: default
  html_document: default
  pdf_document: default
bibliography: Penobscot_Lit.bib
biblio-style: apalike
nocite: '@*'
---
# Technical Report: Habitat Suitability Models for Diadromous Fish in the Northeast United States

## Preface

## Abstract

## Introduction

Diadromous fish species exhibit a remarkable migratory behavior between marine and freshwater environments, playing a pivotal role in the ecological and socio-economic fabric of coastal communities. Their migration enables access to essential resources, identification of suitable spawning grounds, and maintenance of population connectivity, contributing to the overall health and resilience of aquatic ecosystems (Durbin et al., 1979)(Walters et al., 2009). These migratory patterns are categorized into anadromy, involving spawning in fresh or brackish inland water, and catadromy, where reproduction occurs at sea (Limburg & Waldman, 2009). The economic and cultural implications of these migratory movements are significant, as diadromous fish provide essential food resources, support recreational fisheries, and hold cultural importance for coastal and indigenous communities. Understanding and conserving these remarkable migratory species are crucial for sustaining livelihoods, traditions, and cultural heritage associated with coastal communities.

In the Northeastern United States, diadromous fish populations have faced a historic decline over the past century due to dam construction in rivers, overexploitation, and pollution (Brown et al., 2000; Helfman, 2007; Limburg & Waldman, 2009; Hare et al., 2021). Since the enactment of the Endangered Species Act (16 U.S.C. 1531-1544), a total of 47 anadromous fish species or populations have been federally listed as endangered or threatened, underscoring the urgency for conservation efforts (Federally Integrated Species Health (FISH) Act, 2017). However, no federally listed catadromous fish species currently exist. Christensen et al. (2003) revealed a significant decline in the biomass of high-trophic level fishes in the North Atlantic over the past century, highlighting the far-reaching consequences of these population declines on coastal communities.

Effective fisheries management and conservation efforts necessitate an understanding of the habitat requirements and distribution patterns of diadromous fish species (Helfman, 2007). By studying and monitoring the habitats used by these fish, key areas for protection and restoration can be identified, sustainable fishing practices can be implemented, and measures can be taken to mitigate pollution and habitat alteration. Preserving and restoring these habitats are critical steps toward ensuring the recovery and resilience of diadromous fish populations, which, in turn, supports the economic viability and cultural identity of coastal communities.

Although existing HSI models have significantly contributed to understanding habitat preferences and guiding conservation efforts, it is important to acknowledge their limitations. Many of these models, developed based on the best available data and observations at the time, are outdated and often fail to accurately capture the complexity of diadromous fish populations. They are often based on data and observations that group together similar species, such as Atlantic and Shortnose Sturgeon or alewives and blueback herring, thereby overlooking the distinct ecological requirements and behaviors of these species (i.e., Pardue, 1983, add more ). Additionally, models can rely on observation methods that may not be directly applicable to specific ecosystems, such as the U.S. Fish and Wildlife Service Habitat Evaluation Procedures Program, which primarily focuses on developing detailed models for assessing relatively small areas in terrestrial or freshwater environments, can introduce biases and hinder accurate assessments (USFWS 1980a, 1980b, 1981; Able et al., 2020).

Technological advancements such as acoustic telemetry, remote sensing, eDNA analysis, and high-resolution imaging have facilitated the increase in observations, studies, and data for diadromous fish behavior and movement. This wealth of new information has provided valuable insights into specific habitat requirements, migration patterns, and population dynamics of individual species within the diadromous fish community. To reflect these advancements more accurately, it is necessary to develop and refine additional HSI models.  

Refined models incorporating these index approaches will exhibit direct sensitivity to alterations in physical characteristics, ensuring a more precise representation of habitat suitability for each diadromous fish species. This improved accuracy enhances the assessment of the impact of changing environmental conditions on the distribution and abundance of diadromous fish throughout the Northeast United States, facilitating targeted conservation efforts and effective fisheries management strategies. These updated models can also account for the unique ecological characteristics of each species, enabling more precise assessments of their distribution and the identification of critical areas for juvenile and spawning adults. The objectives of this paper are to synthesize the latest life cycle observations, studies, and data on diadromous fish species in the Northeastern United States and develop updated species-specific habitat suitability index (HSI) models for Atlantic Salmon, Atlantic Sturgeon, Alewives, American Eels, American Shad, Atlantic Tomcod, Blueback Herring, Rainbow Smelt, Shortnose Sturgeon, and Striped Bass, considering their unique habitat preferences and ecological needs during the juvenile and spawning adult life stages. By achieving these objectives, this study aims to contribute to a broader understanding of diadromous fish ecology, support effective fisheries management, and inform conservation strategies for the benefit of both the species and the coastal communities dependent on them.

## Methods

## Alewives

Alewives are widely distributed throughout the northeastern United States, inhabiting coastal areas from the Gulf of St. Lawrence in Canada to the mid-Atlantic region of the United States (Collette & Klein-MacPhee, 2003). They are particularly abundant in freshwater rivers and estuaries along the Atlantic coast. Figure x shows the range of alewife habitat throughout the Northeast. Alewife populations can be found in various states, including Maine, New Hampshire, Massachusetts, Rhode Island, Connecticut, New York, New Jersey, and Delaware, among others. They have historically undertaken extensive spawning migrations, ascending rivers and streams from their marine habitat to reach their preferred spawning grounds in freshwater tidal systems (Hare et al., 2021). 

Alewives have experienced significant declines across their distribution range, leading to their designation as a "species of concern" by the U.S. National Marine Fisheries Service in 2006. Furthermore, they have been considered as potential candidates for inclusion in the U.S. Endangered Species List, as highlighted in reports by NOAA NMFS in 2009 and 2013 (NOAA NMFS 2009, 2013).Alewives were once again listed as a candidate species for the Endangered Species List in 2017 (NMFS 2017). The distribution of alewife populations is influenced by several factors, including the availability of suitable spawning habitats, water quality conditions, and the presence of appropriate food resources (Lynch et al., 2015).

### Life cycle overview

The alewife exhibits a complex life cycle characterized by distinct stages and behaviors. The spawning of alewives typically occurs in waves during the spring season, triggered by rising water temperatures and increasing day length (Able et al., 2020; ASMFC 2009). Adult alewives migrate upstream from their marine environment to reach suitable spawning habitats (Bigelow & Schroeder, 1953). Upon arrival, they engage in impressive spawning runs, where large aggregations of alewives gather at spawning grounds.

During the spawning process, male alewives release sperm into the water, while females deposit their adhesive eggs over gravel or other hard-substrate types (Pardue, 1983; Janssen & Luebke, 2004). The adhesive nature of the eggs helps them attach to the substrate, protecting them from being washed away by the current (Brown et al., 2000; Janssen & Brandt, 1980). Both males and females return to the marine environment after spawning (Bigelow & Schroeder, 1953). Although, recent studies support that alewives can make multiple spawning trips upstream during the migratory season (McCartin et al., 2019).

After the spawning process, the eggs of alewives go through an incubation period that typically lasts for several days, as described by Bigelow & Schroeder (1953). Once the eggs hatch, the resulting larvae begin their migration upstream, gradually making their way towards the estuary habitat where they will reside as they grow. This estuary environment serves as a nursery for the juvenile alewife until they eventually migrate out to the sea, as documented by Pardue (1983). Furthermore, Kissil (1974) reported a remarkably low survival rate for the eggs spawned, with only a mere 0.0013% of the 80,000 eggs surviving to leave the spawning grounds as a young fish. Additionally, Kissil's research indicated that, on average, 2.88 fry successfully returned to the sea for each female alewife that entered the spawning grounds.

### Spawning Adult Alewives

#### Habitat Requirements 

Spawning adult alewives exhibit specific preferences and requirements related to various habitat factors. Their annual migration during spawning is energetically demanding, with notable variations in behavior observed. Bigelow & Schroeder (1953) report that adult alewives migrate upstream and refrain from eating until their return downstream to productive tidal habitats. In contrast, Janssen & Brandt (1980) document fasting during the day and extensive feeding at night.

Temperature preferences of alewives during spawning exhibit variations across different studies; however, there is a general consensus that optimal temperatures for successful spawning fall within the range of 12 to 16 degrees Celsius (Brown et al., 2000). Above 29.7 degrees Celsius, spawning success becomes limited (Pardue). Alewives typically migrate into freshwater streams when water temperatures range from 8 to 18 degrees Celsius (Mather et al., 2012). Spawning ceases when water temperatures exceed 27 degrees Celsius (Kissil, 1974), and suitable spawning temperatures broadly span from 10 to 22 degrees Celsius (Tyus, 1974; Pardue, 1983; Collette and Klein-MacPhee, 2002). Alewife eggs can develop within a temperature range of 11 to 28 degrees Celsius (Klauda et al., 1991). It is important to note that temperature plays a critical role in alewife spawning behavior, and deviations from the optimal temperature range can significantly impact spawning success and the timing of migration.

In terms of depth preferences, spawning adult alewives are generally known to favor depths ranging from MLT-10 meters (Brown et al., 2000). However, recent field observations conducted by Mather et al. (2012) have indicated that a significant proportion of alewives can be found in habitats shallower than 2 meters. These findings suggest that alewives are capable of spawning in both shallow and deep water environments, highlighting their adaptability in selecting suitable spawning locations (O'Connell & Angermeier, 1997).

The current understanding of adult alewife spawning behavior indicates a deviance from the traditional assumption that anadromous species exclusively rely on freshwater environments for reproduction. Notably, Brown et al. (2000) emphasizes a heightened preference for habitats with salinity concentrations below 15 psu, while concentrations surpassing 20 psu are deemed unsuitable for spawning adults. This observed variability in spawning behavior suggests that alewives possess the ability to tolerate moderate levels of salinity during the reproductive process. Field studies have documented adult alewives engaging in spawning activities across a diverse array of estuarine habitats, including ponds within coastal systems and pond-like regions within coastal rivers and streams (Pardue 1983; Mullen et al. 1986; Collette and Klein-MacPhee 2002; Walsh et al. 2005). These habitats encompass various environments, such as ponds, oxbows, eddies, backwaters, stream pools, and flooded swamps, and are characterized by their typical attributes of deep and slow water flow (Pardue 1983; Mullen et al. 1986; Collette and Klein-MacPhee 2002; O’Connell and Angermeier 1997; Walsh et al. 2005). Consequently, this revised understanding challenges the longstanding notion that alewives are exclusively obligated to freshwater for spawning purposes, as highlighted by laboratory experiments conducted by DiMaggio et al. (2016) that demonstrate the notable survival rates of alewife embryos at salinities ranging up to and including 30‰.

Flow velocity is a crucial factor influencing the spawning of alewives. O'Connell (1999) proposes that habitat patches with velocity values ranging from 0.06 m/s to 0.16 m/s have a 50% probability of alewife egg presence, while Pardue identifies velocities below 0.3 m/s as suitable habitat for spawning. It should be noted that higher flow velocities have the potential to displace recently spawned eggs from their initial spawning location, as indicated by Able et al.

Previous studies have presented conflicting information regarding the substrate preferences of spawning adult alewives. While some studies suggest that alewives prefer to spawn over hard substrates such as gravel and rock (Pardue, Brown et al., 2000), potentially due to the eggs' better adhesion to such substrates, recent research provides evidence supporting a broader range of substrate utilization by alewives. Recent studies, documented observations of alewives spawning over sandy substrates along with egg presence near hard substrates (Able et al. 2020). Notebly, O’Connell & Angermeier, 1997 also supports the idea that alewives are known to spawn over a range of substrates, including gravel, sand, and other soft substrates.

#### Habitat suitability models

The Alewives Habitat Suitability models, originally developed by Brown et al. (2000) and Pardue (1983), with reliance on similar sources such as Bigelow and Schroeder (1953), possess several limitations that make them inadequate for current applications. Primarily, these models are constructed solely on observations of alewives' daytime behavior, neglecting their significant nocturnal activity patterns. Recent studies have revealed that alewives are primarily active at night, engaging in feeding and exhibiting substantial downstream movement during these nocturnal periods (Janssen & Brandt, 1980; Janssen, 1978; McCartin et al., 2019). Consequently, the exclusive focus on daytime behavior in the existing models fails to capture the true habitat preferences and requirements of alewives, particularly in estuarine and brackish environments.

Furthermore, the current models predominantly consider variables such as temperature, depth, and substrate, while disregarding other crucial factors that significantly influence alewives' habitat selection, including water flow and life stage differences. This limited scope results in incomplete assessments of their habitat suitability. Additionally, inconsistencies and potential inaccuracies arise within the models due to conflicting information regarding the substrate and depth preferences of alewives. These deficiencies undermine the models' effectiveness in predicting the suitability of habitats for alewives.

To address these shortcomings, updated models should encompass a more comprehensive understanding of alewives' behavior, specifically acknowledging their use of estuarine and brackish habitats. These habitats serve as critical areas for alewives, exhibiting relatively high levels of habitat use (McCartin et al., 2019). Incorporating these estuarine and brackish areas into management strategies is of paramount importance to ensure the conservation and successful management of the species. Notably, utilizing estuaries and brackish habitats for spawning may offer energetically favorable conditions for alewives, as it eliminates the need for them to acclimate to complete freshwater environments (DiMaggio et al., 2015). This recognition highlights the significance of incorporating these habitats into conservation efforts and management plans to safeguard the species and support their reproductive success.

In light of the aforementioned limitations, updated suitability indices for spawning adult alewives have been developed to provide a more comprehensive understanding of their habitat preferences and requirements. These indices, presented in Figure 2, incorporate key factors influencing spawning success, including depth, salinity, temperature, flow velocity, and substrate. Suitability index breakpoints and formulas are listed in Table 1.

The depth suitability index, shown in Figure 2a, indicates the preferred depth ranges for spawning adult alewives. Alewives exhibit higher suitability (index of 1) for depths ranging from 2 to 10 meters. Shallower depths below 2 meters and deeper depths exceeding 10 meters have a reduced suitability index of 0.5.

The salinity suitability index reflects the salinity preferences of spawning adult alewives and is presented in Figure 2b. Alewives show a heightened preference for salinity concentrations below 15 psu, resulting in a suitability index of 1. Salinity concentrations exceeding 20 psu are considered unsuitable, with a suitability index of 0.5.

The temperature suitability index, represented in Figure 2c, depicts the optimal temperature range for successful spawning. Alewives exhibit a high suitability index of 1 within the temperature range of 12 to 16 degrees Celsius. Temperatures above 29.7 degrees Celsius have limited suitability, with a suitability index of 0.5.

The flow velocity suitability index, shown in Figure 2d, indicates the preferred flow velocity range for spawning. Habitat patches with flow velocities ranging from 0.06 to 0.16 m/s have a high suitability index of 1. Velocities below 0.3 m/s are also suitable, albeit with a reduced suitability index of 0.5.

Lastly, the substrate suitability index assesses the preference for different substrate types and is presented in Figure 2e. Alewives show a higher suitability index of 0.8 for hard substrates such as gravel or rock. Sandy substrates have a slightly lower suitability index of 0.6, while other soft substrates exhibit a suitability index of 0.7.

```{r, echo= FALSE, include=TRUE, fig.align = "center", fig.cap="Habitat Range of Alewive in the Northeast United States", out.width="80%"}
# Required packages
#library(ggplot2)
#library(sf)

# Read the alewife habitat data (replace "alewife_habitat.shp" with the actual file name and path)
#alewife_habitat <- st_read("alewife_habitat.shp")

# Plot the map
#ggplot() +
#  geom_sf(data = alewife_habitat, fill = "lightblue", color = "black") +
#  coord_sf() +
#  labs(title = "Alewife Habitat Distribution",
#       subtitle = "Northeastern United States",
#       x = "Longitude", y = "Latitude")
```

```{r, echo= FALSE, include=TRUE, fig.align = "center", fig.cap="Spawning Adult Alewife Suitability Breakpoints", out.width="80%"}
library(knitr)

# Create the data for the table
index <- c("a. Depth", "b. Salinity", "c. Temperature", "d. Flow Velocity", "e. Substrate")
breakpoints <- c("Depth ≤ 2 m: 0.5\nDepth > 2 m: 1",
                 "Salinity < 15 psu: 1\nSalinity ≤ 20 psu: 0.5",
                 "Temperature 12-16°C: 1\nTemperature > 16°C & ≤ 29.7°C: 0.5",
                 "Velocity 0.06-0.16 m/s: 1\nVelocity < 0.3 m/s: 0.5",
                 "Hard (gravel or rock): 0.8\nSandy: 0.6\nSoft (other substrates): 0.7\nSuspended sediments: 0.9")

# Create a data frame
df <- data.frame(Index = index, Breakpoints = breakpoints)

# Print the table using kable
kable(df, format = "markdown")
```

```{r, echo= FALSE, include=TRUE, fig.align = "center", fig.cap="Spawning Adult Alewife Suitability Indices", out.width="80%"}
par(mfrow=c(2, 3))

# Depth Suitability Index
depth_values <- seq(0, 30, length.out = 100)
depth_suitability <- ifelse(depth_values <= 2, 0.5, ifelse(depth_values <= 10, 1, 0))
plot(depth_values, depth_suitability, type = "l", xlab = "Depth (m)", ylab = "Suitability Index")
title("Depth Suitability Index")

# Salinity Suitability Index
salinity_values <- seq(0, 30, length.out = 100)
salinity_suitability <- ifelse(salinity_values < 15, 1, ifelse(salinity_values <= 20, 0.5, 0))
plot(salinity_values, salinity_suitability, type = "l", xlab = "Salinity (psu)", ylab = "Suitability Index")
title("Salinity Suitability Index")

# Temperature Suitability Index
temperature_values <- seq(0, 30, length.out = 100)
temperature_suitability <- ifelse(temperature_values >= 12 & temperature_values <= 16, 1, ifelse(temperature_values > 16 & temperature_values <= 29.7, 0.5, 0))
plot(temperature_values, temperature_suitability, type = "l", xlab = "Temperature (°C)", ylab = "Suitability Index")
title("Temperature Suitability Index")

# Flow Velocity Suitability Index
velocity_values <- seq(0, 1, length.out = 100)
velocity_suitability <- ifelse(velocity_values >= 0.06 & velocity_values <= 0.16, 1, ifelse(velocity_values < 0.3, 0.5, 0))
plot(velocity_values, velocity_suitability, type = "l", xlab = "Flow Velocity (m/s)", ylab = "Suitability Index")
title("Flow Velocity Suitability Index")

# Substrate Suitability Index
substrate <- c("Hard (gravel or rock)", "Sandy", "Soft (other substrates)")
substrate_suitability <- c(0.8, 0.6, 0.7)
barplot(substrate_suitability, names.arg = substrate, xlab = "Substrate", ylab = "Suitability Index")
title("Substrate Suitability Index")
```

- juvenile
Janssen & Luebke, 2004 describes a significant observations of juvenile herring over rocky substrates.
#### Spawning
#### Juvenile
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

## American Eels
- description of species
- map of geographic range
- description of northeast distribution

### Life cycle overview
### Habitat requirements
- spawning
- juvenile

### Habitat suitability models
#### spawning
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

#### juvenile
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

## American Shad
- description of species
- map of geographic range
- description of northeast distribution

### Life cycle overview
### Habitat requirements
- spawning
- juvenile

### Habitat suitability models
#### Spawning
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

#### Juvenile
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

## Atlantic Salmon
- description of species
- map of geographic range
- description of northeast distribution

### Life cycle overview
### Habitat requirements
- spawning
- juvenile

### Habitat suitability models
#### Spawning
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

#### Juvenile
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

## Atlantic Sturgeon
- description of species
- map of geographic range
- description of northeast distribution

### Life cycle overview
### Habitat requirements
- spawning
- juvenile

### Habitat suitability models
#### Spawning
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

#### Juvenile
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

## Atlantic Tomcod
- description of species
- map of geographic range
- description of northeast distribution

### Life cycle overview

### Habitat requirements
- spawning
- juvenile

### Habitat suitability models
#### Spawning
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

#### Juvenile
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

## Blueback Herring
- description of species
- map of geographic range
- description of northeast distribution

### Life cycle overview
### Habitat requirements
- spawning
- juvenile

### Habitat suitability models
#### Spawning
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

#### Juvenile
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

## Rainbow Smelt
- description of species
- map of geographic range
- description of northeast distribution

### Life cycle overview
### Habitat requirements
- spawning
- juvenile

### Habitat suitability models
#### Spawning
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

#### Juvenile
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

## Shortnose Sturgeon
- description of species
- map of geographic range
- description of northeast distribution

### Life cycle overview
### Habitat requirements
- spawning
- juvenile

### Habitat suitability models
#### Spawning
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

#### Juvenile
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

## Striped Bass
- description of species
- map of geographic range
- description of northeast distribution

### Life cycle overview
### Habitat requirements
- spawning
- juvenile

### Habitat suitability models
#### Spawning
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

#### Juvenile
##### Depth
##### Salinity
##### Temperature
##### Flow Velocity
##### Sediment Composition

## Synthesis and Discussion
- Comparison of habitat suitability models across species
- Implications for fisheries management and conservation
- Future directions and potential improvements for habitat suitability models

## Conclusion
- Summary of key findings
- Importance of habitat suitability models for diadromous fish management
- Final remarks and call to action