---
title: "Blueback Herring Habitat Suitability Model"
author: "Vanessa Mahan"
date: "2023-04-19"
output: pdf_document
---
# Blueback Herring

## Spawning Adult Habitat Suitability

```{r echo=FALSE, fig.width=14, fig.height=10, fig.cap="Suitability index curves for Blueback Herring Spawning Habitat"}
#Create data frame with freshwater tidal submodel
bluebackherring.spawning <- data.frame(matrix(NA,nrow=7,ncol=12))
colnames(bluebackherring.spawning) <- c("depth","depth.SIV","salinity", "salinity.SIV", "temp.C", "temp.C.SIV","velocity","velocity.SIV","silt.clay.per","silt.clay.SIV","tubidity","turbidity.SIV")

bluebackherring.spawning$depth <- c(0,1,10,20,50,NA,NA)
bluebackherring.spawning$depth.SIV <- c(0,1,1,0.1,0.1,NA,NA)
bluebackherring.spawning$salinity <- c(0.5,5,10,15,20,35,NA)
bluebackherring.spawning$salinity.SIV <- c(1,1,1,0.5,0,0,NA)
bluebackherring.spawning$temp.C <- c(0,4,7,11,28,30,45)
bluebackherring.spawning$temp.C.SIV <- c(0.1,0.5,0.5,1,1,0,0)
bluebackherring.spawning$velocity <- c(0,0.05,0.3,1,5,NA,NA)
bluebackherring.spawning$velocity.SIV <- c(0,1,1,0,0,NA,NA)
bluebackherring.spawning$silt.clay.per <- c(0,0.2,0.4,0.8,1,NA,NA)
bluebackherring.spawning$silt.clay.SIV <- c(1,1,0.5,0,0,NA,NA)
bluebackherring.spawning$tubidity <- c(0,0.2,0.4,0.8,1,NA,NA)
bluebackherring.spawning$turbidity.SIV <- c(1,1,0.5,0,0,NA,NA)

##########
#Create summary figure for fresh.tid submodel
par(mfrow=c(2,3), cex=1.0)

  #Depth
  plot(bluebackherring.spawning$depth, bluebackherring.spawning$depth.SIV, xlim=c(0,100), ylim=c(0,1), type="l", xlab="Depth (meters)", ylab="Suitability Index (depth)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.spawning$depth, bluebackherring.spawning$depth.SIV, pch=19, cex=2)
  lines(bluebackherring.spawning$depth, bluebackherring.spawning$depth.SIV, lwd=3)
  
  #salinity.
  plot(bluebackherring.spawning$salinity, bluebackherring.spawning$salinity.SIV, xlim=c(0,35), ylim=c(0,1), type="l",
       xlab="Salinity (psu)", ylab="Suitability Index (salinity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.spawning$salinity, bluebackherring.spawning$salinity.SIV, pch=19, cex=2)
  lines(bluebackherring.spawning$salinity, bluebackherring.spawning$salinity.SIV, lwd=3)
  
  #temperature
  plot(bluebackherring.spawning$temp.C, bluebackherring.spawning$temp.C.SIV, xlim=c(0,50), ylim=c(0,1), type="l",
       xlab="Temperature (Celcius)", ylab="Suitability Index (temperature)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.spawning$temp.C, bluebackherring.spawning$temp.C.SIV, pch=19, cex=2)
  lines(bluebackherring.spawning$temp.C, bluebackherring.spawning$temp.C.SIV, lwd=3)
  
  #velocity
  plot(bluebackherring.spawning$velocity, bluebackherring.spawning$velocity.SIV, xlim=c(0,8), ylim=c(0,1), type="l",
       xlab="Velocity (m/s)", ylab="Suitability Index (velocity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.spawning$velocity, bluebackherring.spawning$velocity.SIV, pch=19, cex=2)
  lines(bluebackherring.spawning$velocity, bluebackherring.spawning$velocity.SIV, lwd=3)
  
  #silt/clay percentage
  plot(bluebackherring.spawning$silt.clay.per, bluebackherring.spawning$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.spawning$silt.clay.per, bluebackherring.spawning$silt.clay.SIV, pch=19, cex=2)
  lines(bluebackherring.spawning$silt.clay.per, bluebackherring.spawning$silt.clay.SIV, lwd=3)
  
   #turbidity
  plot(bluebackherring.spawning$silt.clay.per, bluebackherring.spawning$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.spawning$silt.clay.per, bluebackherring.spawning$silt.clay.SIV, pch=19, cex=2)
  lines(bluebackherring.spawning$silt.clay.per, bluebackherring.spawning$silt.clay.SIV, lwd=3)
```

#### Depth
According to (O’Connell & Angermeier, 1997), Alewives spawn in both shallow and deep water.

#### Salinity

#### Temperature

Pardue decribes optimal spawning temperature for Blueback Herring between 20-24 degrees celcius. In contrast Tommasi et al 2015 describes optimal spawning temperatures around 11 degrees to maximize abundance of juvenile blueback herring. O'connell 1999 mentions that habitat patches with a temperature of 16.8 degrees celcius to have a probability of 0.5 for Blueback Herring egg presence. The paper also mentions high egg densities from 11-28 degrees celcius. Hatching success in impaired above 28 degrees (Pardue & O'Connell). Blueback Herring eggs were found in 1996 between 11.4-23  and in 1997 10.98C to 20.88C degrees Celcius (Walsh et al., 2005).Temperatures from 14-26 are suitable for alewife eggs (Klauda et al. 1991).

#### Flow Velocity

#### Sediment Composition
Prefer to spawn over hard substrate (O’Connell & Angermeier, 1997).

#### Turbity 

## Juvenile Habitat
```{r echo=FALSE, fig.width=14, fig.height=10, fig.cap="Suitability index curves for Juvenile Blueback Herring Habitat"}
#Create data frame with freshwater tidal submodel
bluebackherring.juv <- data.frame(matrix(NA,nrow=6,ncol=12))
colnames(bluebackherring.juv) <- c("depth","depth.SIV","salinity", "salinity.SIV", "temp.C", "temp.C.SIV","velocity","velocity.SIV","silt.clay.per","silt.clay.SIV","tubidity","turbidity.SIV")

bluebackherring.juv$depth <- c(0,1,10,20,50,NA)
bluebackherring.juv$depth.SIV <- c(0,1,1,0,0,NA)
bluebackherring.juv$salinity <- c(0.5,5,15,25,30,NA)
bluebackherring.juv$salinity.SIV <- c(0.5,1,1,0.5,0.1,NA)
bluebackherring.juv$temp.C <- c(0,2,4,12,14,16)
bluebackherring.juv$temp.C.SIV <- c(0,0.5,1,1,0.5,0)
bluebackherring.juv$velocity <- c(0,0.05,0.3,1,5,NA)
bluebackherring.juv$velocity.SIV <- c(0,1,1,0,0,NA)
bluebackherring.juv$silt.clay.per <- c(0,0.2,0.4,0.8,1,NA)
bluebackherring.juv$silt.clay.SIV <- c(1,1,0.5,0,0,NA)
bluebackherring.juv$tubidity <- c(0,0.2,0.4,0.8,1,NA)
bluebackherring.juv$turbidity.SIV <- c(1,1,0.5,0,0,NA)

##########
#Create summary figure for fresh.tid submodel
par(mfrow=c(2,3), cex=1.0)

  #Depth
  plot(bluebackherring.juv$depth, bluebackherring.juv$depth.SIV, xlim=c(0,100), ylim=c(0,1), type="l", xlab="Depth (meters)", ylab="Suitability Index (depth)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.juv$depth, bluebackherring.juv$depth.SIV, pch=19, cex=2)
  lines(bluebackherring.juv$depth, bluebackherring.juv$depth.SIV, lwd=3)
  
  #salinity.
  plot(bluebackherring.juv$salinity, bluebackherring.juv$salinity.SIV, xlim=c(0,35), ylim=c(0,1), type="l",
       xlab="Salinity (psu)", ylab="Suitability Index (salinity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.juv$salinity, bluebackherring.juv$salinity.SIV, pch=19, cex=2)
  lines(bluebackherring.juv$salinity, bluebackherring.juv$salinity.SIV, lwd=3)
  
  #temperature
  plot(bluebackherring.juv$temp.C, bluebackherring.juv$temp.C.SIV, xlim=c(0,50), ylim=c(0,1), type="l",
       xlab="Temperature (Celcius)", ylab="Suitability Index (temperature)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.juv$temp.C, bluebackherring.juv$temp.C.SIV, pch=19, cex=2)
  lines(bluebackherring.juv$temp.C, bluebackherring.juv$temp.C.SIV, lwd=3)
  
  #velocity
  plot(bluebackherring.juv$velocity, bluebackherring.juv$velocity.SIV, xlim=c(0,8), ylim=c(0,1), type="l",
       xlab="Velocity (m/s)", ylab="Suitability Index (velocity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.juv$velocity, bluebackherring.juv$velocity.SIV, pch=19, cex=2)
  lines(bluebackherring.juv$velocity, bluebackherring.juv$velocity.SIV, lwd=3)
  
  #silt/clay percentage
  plot(bluebackherring.juv$silt.clay.per, bluebackherring.juv$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.juv$silt.clay.per, bluebackherring.juv$silt.clay.SIV, pch=19, cex=2)
  lines(bluebackherring.juv$silt.clay.per, bluebackherring.juv$silt.clay.SIV, lwd=3)
  
   #turbidity
  plot(bluebackherring.juv$silt.clay.per, bluebackherring.juv$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(bluebackherring.juv$silt.clay.per, bluebackherring.juv$silt.clay.SIV, pch=19, cex=2)
  lines(bluebackherring.juv$silt.clay.per, bluebackherring.juv$silt.clay.SIV, lwd=3)
```

#### Depth


#### Salinity


#### Temperature

Tommasi Et al 2015 describes optimal nursery habitat with an optimal temperature range between 20-24 degrees celcius with a total range of suitability from 16-26 degrees celcuis. Temperature also verified in where 100% survival is shown near 30 psu and decreased survival is experiences around 15 psu further at 0 psu (DiMaggio 2015).

#### Flow Velocity

#### Sediment Composition

#### Turbity 

## References