---
title: "Atlantic Tomcod Habitat Suitability Model"
author: "Vanessa Mahan"
date: "2023-04-19"
output: pdf_document
---
# Atlantic Tomcod


## Spawning Adult Habitat Suitability

```{r echo=FALSE, fig.width=14, fig.height=10, fig.cap="Suitability index curves for Spawning Adult Atlantic Tomcod Habitat"}
#Create data frame with freshwater tidal submodel
atlantictomcod.spawning <- data.frame(matrix(NA,nrow=7,ncol=12))
colnames(atlantictomcod.spawning) <- c("depth","depth.SIV","salinity", "salinity.SIV", "temp.C", "temp.C.SIV","velocity","velocity.SIV","silt.clay.per","silt.clay.SIV","tubidity","turbidity.SIV")

atlantictomcod.spawning$depth <- c(0,1,10,20,50,NA,NA)
atlantictomcod.spawning$depth.SIV <- c(0,1,1,0.1,0.1,NA,NA)
atlantictomcod.spawning$salinity <- c(0.5,5,10,15,20,35,NA)
atlantictomcod.spawning$salinity.SIV <- c(1,1,1,0.5,0,0,NA)
atlantictomcod.spawning$temp.C <- c(0,2,8,10,16,18,26)
atlantictomcod.spawning$temp.C.SIV <- c(0,0.1,0.5,1,1,0.5,0.1)
atlantictomcod.spawning$velocity <- c(0,0.05,0.3,1,5,NA,NA)
atlantictomcod.spawning$velocity.SIV <- c(0,1,1,0,0,NA,NA)
atlantictomcod.spawning$silt.clay.per <- c(0,0.2,0.4,0.8,1,NA,NA)
atlantictomcod.spawning$silt.clay.SIV <- c(1,1,0.5,0,0,NA,NA)
atlantictomcod.spawning$tubidity <- c(0,0.2,0.4,0.8,1,NA,NA)
atlantictomcod.spawning$turbidity.SIV <- c(1,1,0.5,0,0,NA,NA)

##########
#Create summary figure for fresh.tid submodel
par(mfrow=c(2,3), cex=1.0)

  #Depth
  plot(atlantictomcod.spawning$depth, atlantictomcod.spawning$depth.SIV, xlim=c(0,100), ylim=c(0,1), type="l", xlab="Depth (meters)", ylab="Suitability Index (depth)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.spawning$depth, atlantictomcod.spawning$depth.SIV, pch=19, cex=2)
  lines(atlantictomcod.spawning$depth, atlantictomcod.spawning$depth.SIV, lwd=3)
  
  #salinity.
  plot(atlantictomcod.spawning$salinity, atlantictomcod.spawning$salinity.SIV, xlim=c(0,35), ylim=c(0,1), type="l",
       xlab="Salinity (psu)", ylab="Suitability Index (salinity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.spawning$salinity, atlantictomcod.spawning$salinity.SIV, pch=19, cex=2)
  lines(atlantictomcod.spawning$salinity, atlantictomcod.spawning$salinity.SIV, lwd=3)
  
  #temperature
  plot(atlantictomcod.spawning$temp.C, atlantictomcod.spawning$temp.C.SIV, xlim=c(0,50), ylim=c(0,1), type="l",
       xlab="Temperature (Celcius)", ylab="Suitability Index (temperature)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.spawning$temp.C, atlantictomcod.spawning$temp.C.SIV, pch=19, cex=2)
  lines(atlantictomcod.spawning$temp.C, atlantictomcod.spawning$temp.C.SIV, lwd=3)
  
  #velocity
  plot(atlantictomcod.spawning$velocity, atlantictomcod.spawning$velocity.SIV, xlim=c(0,8), ylim=c(0,1), type="l",
       xlab="Velocity (m/s)", ylab="Suitability Index (velocity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.spawning$velocity, atlantictomcod.spawning$velocity.SIV, pch=19, cex=2)
  lines(atlantictomcod.spawning$velocity, atlantictomcod.spawning$velocity.SIV, lwd=3)
  
  #silt/clay percentage
  plot(atlantictomcod.spawning$silt.clay.per, atlantictomcod.spawning$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.spawning$silt.clay.per, atlantictomcod.spawning$silt.clay.SIV, pch=19, cex=2)
  lines(atlantictomcod.spawning$silt.clay.per, atlantictomcod.spawning$silt.clay.SIV, lwd=3)
  
   #turbidity
  plot(atlantictomcod.spawning$silt.clay.per, atlantictomcod.spawning$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.spawning$silt.clay.per, atlantictomcod.spawning$silt.clay.SIV, pch=19, cex=2)
  lines(atlantictomcod.spawning$silt.clay.per, atlantictomcod.spawning$silt.clay.SIV, lwd=3)
```

#### Depth

According to Brown et al (2000), spawning adult Altantic Tomcod have a depth preference for MLT-10 meters. There is also a very low preference for habitat greater than 20 meters in depth. 

#### Salinity
Brown et al (2000) also describes a higher tolerance for salinity with high preference for habitat with salinity concentrations of <10 psu. Salinty concentrations greater than 15 are unsuitable for spawning adults.

#### Temperature
Brown et al (2000) decribes a wide range of temperature preferences with optimal temperatures at 10-16 degrees celcius. The total range of temperatures suitable for spawning alewives is 2-26 degrees Celsius.  8&18 have 0.5

#### Flow Velocity

#### Sediment Composition

#### Turbity 

## Juvenile Habitat
```{r echo=FALSE, fig.width=14, fig.height=10, fig.cap="Suitability index curves for Juvenile Atlantic Salmon Habitat"}
#Create data frame with freshwater tidal submodel
atlantictomcod.juv <- data.frame(matrix(NA,nrow=6,ncol=12))
colnames(atlantictomcod.juv) <- c("depth","depth.SIV","salinity", "salinity.SIV", "temp.C", "temp.C.SIV","velocity","velocity.SIV","silt.clay.per","silt.clay.SIV","tubidity","turbidity.SIV")

atlantictomcod.juv$depth <- c(0,1,10,20,50,NA)
atlantictomcod.juv$depth.SIV <- c(0,1,1,0,0,NA)
atlantictomcod.juv$salinity <- c(0.5,5,15,25,30,NA)
atlantictomcod.juv$salinity.SIV <- c(0.5,1,1,0.5,0.1,NA)
atlantictomcod.juv$temp.C <- c(0,2,4,12,14,16)
atlantictomcod.juv$temp.C.SIV <- c(0,0.5,1,1,0.5,0)
atlantictomcod.juv$velocity <- c(0,0.05,0.3,1,5,NA)
atlantictomcod.juv$velocity.SIV <- c(0,1,1,0,0,NA)
atlantictomcod.juv$silt.clay.per <- c(0,0.2,0.4,0.8,1,NA)
atlantictomcod.juv$silt.clay.SIV <- c(1,1,0.5,0,0,NA)
atlantictomcod.juv$tubidity <- c(0,0.2,0.4,0.8,1,NA)
atlantictomcod.juv$turbidity.SIV <- c(1,1,0.5,0,0,NA)

##########
#Create summary figure for fresh.tid submodel
par(mfrow=c(2,3), cex=1.0)

  #Depth
  plot(atlantictomcod.juv$depth, atlantictomcod.juv$depth.SIV, xlim=c(0,100), ylim=c(0,1), type="l", xlab="Depth (meters)", ylab="Suitability Index (depth)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.juv$depth, atlantictomcod.juv$depth.SIV, pch=19, cex=2)
  lines(atlantictomcod.juv$depth, atlantictomcod.juv$depth.SIV, lwd=3)
  
  #salinity.
  plot(atlantictomcod.juv$salinity, atlantictomcod.juv$salinity.SIV, xlim=c(0,35), ylim=c(0,1), type="l",
       xlab="Salinity (psu)", ylab="Suitability Index (salinity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.juv$salinity, atlantictomcod.juv$salinity.SIV, pch=19, cex=2)
  lines(atlantictomcod.juv$salinity, atlantictomcod.juv$salinity.SIV, lwd=3)
  
  #temperature
  plot(atlantictomcod.juv$temp.C, atlantictomcod.juv$temp.C.SIV, xlim=c(0,50), ylim=c(0,1), type="l",
       xlab="Temperature (Celcius)", ylab="Suitability Index (temperature)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.juv$temp.C, atlantictomcod.juv$temp.C.SIV, pch=19, cex=2)
  lines(atlantictomcod.juv$temp.C, atlantictomcod.juv$temp.C.SIV, lwd=3)
  
  #velocity
  plot(atlantictomcod.juv$velocity, atlantictomcod.juv$velocity.SIV, xlim=c(0,8), ylim=c(0,1), type="l",
       xlab="Velocity (m/s)", ylab="Suitability Index (velocity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.juv$velocity, atlantictomcod.juv$velocity.SIV, pch=19, cex=2)
  lines(atlantictomcod.juv$velocity, atlantictomcod.juv$velocity.SIV, lwd=3)
  
  #silt/clay percentage
  plot(atlantictomcod.juv$silt.clay.per, atlantictomcod.juv$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.juv$silt.clay.per, atlantictomcod.juv$silt.clay.SIV, pch=19, cex=2)
  lines(atlantictomcod.juv$silt.clay.per, atlantictomcod.juv$silt.clay.SIV, lwd=3)
  
   #turbidity
  plot(atlantictomcod.juv$silt.clay.per, atlantictomcod.juv$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(atlantictomcod.juv$silt.clay.per, atlantictomcod.juv$silt.clay.SIV, pch=19, cex=2)
  lines(atlantictomcod.juv$silt.clay.per, atlantictomcod.juv$silt.clay.SIV, lwd=3)
```

#### Depth
Depth preference for juvenile atlantic tomcod was slightly different than spawning adults with a preference of MLT-10 meters with no habitat suitability past 10 meters (Brown et al (2000)).

#### Salinity
Salinity preferences are greater for juvenile atlantic tommcod than spawning adults. Juvenile tomcod were listed to have a high preference for habitat with salinity concentrations <15 psu but show some suitability for estuary environments up to <30 psu (Brown et al (2000)).

#### Temperature
Temperature preferences for juvenile atlantic tomcod is a wider range than spawning adults. Juvenile atlantic tomcod have a preference for temperatures from 0-16 degrees celcius with optimal suitability from 4-12 degrees Celsius (Brown et al (2000)).

#### Flow Velocity

#### Sediment Composition

#### Turbity 

## References