---
title: "Atlantic Salmon Spawning Suitability Indices"
author: "Vanessa Mahan"
date: "2023-04-01"
output: html_document
---
Atlantic Salmon have specific habitat preferences for spawning, and several factors such as depth, salinity, temperature, minimum flow velocity, and silt/clay percentage can influence their choice of spawning location.

Depth: Atlantic Salmon prefer water depths of 30-60 cm (12-24 inches) for spawning. This depth allows for good oxygenation and provides a suitable location for the female to create a nest, or redd.
```{r}
#depth_range <- c(0.305,0.61)#in meters
depth <- (0.305+1)/2 #average between min & max
depth_suitability_curve <- function(depth) {
  # Define the optimal range for depth
  optimal_depth_range <- c(0.3, 1) # meters

  # Define the lower and upper thresholds for depth
  lower_threshold <- 0 # meters
  upper_threshold <- 1.0 # meters

  # Compute the habitat suitability score based on depth
  if (depth < lower_threshold | depth > upper_threshold) {
    suitability_score <- 0
  } else if (depth >= optimal_depth_range[1] & depth <= optimal_depth_range[2]) {
    suitability_score <- 1
  } else if (depth < optimal_depth_range[1]) {
    suitability_score <- (depth - lower_threshold) / (optimal_depth_range[1] - lower_threshold)
  } else if (depth > optimal_depth_range[2]) {
    suitability_score <- (upper_threshold - depth) / (upper_threshold - optimal_depth_range[2])
  }

  return(suitability_score)
}
plot_depth_suitability_curve <- function() {
  # Define the range of depths to plot
  depths <- seq(0, 20, by = 0.01)

  # Compute the habitat suitability score for each depth value
  suitability_scores <- sapply(depths, depth_suitability_curve)

  # Plot the suitability scores as a curve
  plot(depths, suitability_scores, type = "l", xlab = "Depth (meters)", ylab = "Habitat Suitability", ylim=c(0,1),
       main = "Habitat Suitability Curve for Depth Parameter")
}
plot_depth_suitability_curve()
```

Salinity: Atlantic Salmon are anadromous, meaning they spend most of their adult lives at sea but return to freshwater rivers to spawn. As such, they require freshwater for spawning and avoid areas with high salinity.
```{r}
#optimal Salinity range from 0.0psu-0.5psu
Salinity <- (0+0.5)/2 #average between min & max
salinity_suitability_curve <- function(Salinity,optimal_salinity_range) {
  # Define the optimal range for Salinity
  optimal_salinity_range <- c(0, 0.5) #psu

  # Define the lower and upper thresholds for Salinity
  lower_threshold <- 0 # psu
  upper_threshold <- 1 # psu

  # Compute the habitat suitability score based on Salinity
  if (Salinity < lower_threshold | Salinity > upper_threshold) {
    suitability_score <- 0
  } else if (Salinity >= optimal_salinity_range[1] & Salinity <= optimal_salinity_range[2]) {
    suitability_score <- 1
  } else if (Salinity < optimal_salinity_range[1]) {
    suitability_score <- (Salinity - lower_threshold) / (optimal_salinity_range[1] - lower_threshold)
  } else if (Salinity > optimal_salinity_range[2]) {
    suitability_score <- (upper_threshold - Salinity) / (upper_threshold - optimal_salinity_range[2])
  }

  return(suitability_score)
}
plot_salinity_suitability_curve <- function() {
  # Define the range of Salinities to plot
  Salinities <- seq(0, 30, by = 0.01)

  # Compute the habitat suitability score for each depth value
  suitability_scores <- sapply(Salinities, salinity_suitability_curve)

  # Plot the suitability scores as a curve
  plot(Salinities, suitability_scores, type = "l", xlab = "Salinity (psu)", ylab = "Habitat Suitability", ylim=c(0,1),
       main = "Habitat Suitability Curve for Salinity Parameter")
}
plot_salinity_suitability_curve()
```

Temperature: Atlantic Salmon prefer spawning locations with cool water temperatures between 6-15°C (43-59°F). This temperature range allows for optimal egg development and survival.
```{r}
#optimal temperature range from 0.0psu-0.5psu
temperature <- (6+15)/2 #average between min & max
temperature_suitability_curve <- function(temperature,optimal_temp_range) {
  # Define the optimal range for temperature
  optimal_temp_range <- c(6,15) #celcius

  # Define the lower and upper thresholds for temperature
  lower_threshold <- 0 # psu
  upper_threshold <- 20 # psu

  # Compute the habitat suitability score based on temperature
  if (temperature < lower_threshold | temperature > upper_threshold) {
    suitability_score <- 0
  } else if (temperature >= optimal_temp_range[1] & temperature <= optimal_temp_range[2]) {
    suitability_score <- 1
  } else if (temperature < optimal_temp_range[1]) {
    suitability_score <- (temperature - lower_threshold) / (optimal_temp_range[1] - lower_threshold)
  } else if (temperature > optimal_temp_range[2]) {
    suitability_score <- (upper_threshold - temperature) / (upper_threshold - optimal_temp_range[2])
  }

  return(suitability_score)
}
plot_temperature_suitability_curve <- function() {
  # Define the range of temperature to plot
  temps <- seq(0, 30, by = 0.01)

  # Compute the habitat suitability score for each depth value
  suitability_scores <- sapply(temps, temperature_suitability_curve)

  # Plot the suitability scores as a curve
  plot(temps, suitability_scores, type = "l", xlab = "Temperature (Celcius)", ylab = "Habitat Suitability", ylim=c(0,1),
       main = "Habitat Suitability Curve for Temperature Parameter")
}
plot_temperature_suitability_curve()
```

Minimum flow velocity: Atlantic Salmon require a minimum flow velocity of 0.3-0.9 m/s (1-3 ft/s) for spawning. This flow velocity is necessary to keep the eggs and newly hatched fry oxygenated and prevent sediment from settling on them.
```{r}
#depth_range <- c(0.305,0.61)#in meters
flow_velocity <- (0.3-0.9)/2 #average between min & max
flow_velocity_suitability <- function(flow_velocity) {
  # Define the optimal range for depth
  optimal_velocity_range <- c(0.3, 0.9) # meters/second

  # Define the lower and upper thresholds for depth
  lower_threshold <- 0.1 # meters/second
  upper_threshold <- 1.0 # meters/second

  # Compute the habitat suitability score based on depth
  if (flow_velocity < lower_threshold | flow_velocity > upper_threshold) {
    suitability_score <- 0
  } else if (flow_velocity >= optimal_velocity_range[1] & flow_velocity <= optimal_velocity_range[2]) {
    suitability_score <- 1
  } else if (flow_velocity < optimal_velocity_range[1]) {
    suitability_score <- (flow_velocity - lower_threshold) / (optimal_velocity_range[1] - lower_threshold)
  } else if (flow_velocity > optimal_velocity_range[2]) {
    suitability_score <- (upper_threshold - flow_velocity) / (upper_threshold - optimal_velocity_range[2])
  }

  return(suitability_score)
}
plot_Velocity_suitability_curve <- function() {
  # Define the range of depths to plot
  velocities <- seq(0,3, by = 0.01)

  # Compute the habitat suitability score for each depth value
  suitability_scores <- sapply(velocities, flow_velocity_suitability)

  # Plot the suitability scores as a curve
  plot(velocities, suitability_scores, type = "l", xlab = "Velocities (meters/second)", ylab = "Habitat Suitability", ylim=c(0,1),
       main = "Habitat Suitability Curve for Flow Velocity Parameter")
}
plot_Velocity_suitability_curve()
```

Silt/clay percentage: Atlantic Salmon prefer spawning locations with clean, well-oxygenated gravels or cobble substrates, with low silt and clay percentages. High silt and clay percentages can clog the spaces between rocks, reducing water flow and oxygenation, which can negatively affect egg survival.
```{r}
#depth_range <- c(0.305,0.61)#in meters
SiltClay_per <- (0-0.1)/2 #average between min & max
siltclay_suitability <- function(SiltClay_per) {
  # Define the optimal range for depth
  optimal_siltclay_range <- c(0, 0.1) # percent

  # Define the lower and upper thresholds for silt clay percentage
  lower_threshold <- 0 # meters/second
  upper_threshold <- 1 # meters/second

  # Compute the habitat suitability score based on depth
  if (SiltClay_per < lower_threshold | SiltClay_per > upper_threshold) {
    suitability_score <- 0
  } else if (SiltClay_per >= optimal_siltclay_range[1] & SiltClay_per <= optimal_siltclay_range[2]) {
    suitability_score <- 1
  } else if (SiltClay_per < optimal_siltclay_range[1]) {
    suitability_score <- (SiltClay_per - lower_threshold) / (optimal_siltclay_range[1] - lower_threshold)
  } else if (SiltClay_per > optimal_siltclay_range[2]) {
    suitability_score <- (upper_threshold - SiltClay_per) / (upper_threshold - optimal_siltclay_range[2])
  }

  return(suitability_score)
}
plot_SiltClay_suitability_curve <- function() {
  # Define the range of depths to plot
  siltclay_percentile <- seq(0,1, by = 0.01)

  # Compute the habitat suitability score for each depth value
  suitability_scores <- sapply(siltclay_percentile, siltclay_suitability)

  # Plot the suitability scores as a curve
  plot(siltclay_percentile, suitability_scores, type = "l", xlab = "Percent of Silt/Clay in Sediment (%)", ylab = "Habitat Suitability", ylim=c(0,1),
       main = "Habitat Suitability Curve for Sediment Parameter")
}
plot_SiltClay_suitability_curve()
```

Overall, Atlantic Salmon are very sensitive to changes in their environment and require specific habitat conditions for successful spawning. Understanding and protecting these habitat preferences is crucial for the conservation and recovery of Atlantic Salmon populations.