---
title: "Alewife Habitat Suitability Model"
author: "Vanessa Mahan"
date: "2023-04-19"
output: pdf_document
---
# Alewife

## Spawning Adult Habitat Suitability

```{r echo=FALSE, fig.width=14, fig.height=10, fig.cap="Suitability index curves for Spawning Alewife Spawning Habitat"}
#Create data frame with freshwater tidal submodel
alewife.spawning <- data.frame(matrix(NA,nrow=5,ncol=12))
colnames(alewife.spawning) <- c("depth","depth.SIV","salinity", "salinity.SIV", "temp.C", "temp.C.SIV","velocity","velocity.SIV","silt.clay.per","silt.clay.SIV","tubidity","turbidity.SIV")

alewife.spawning$depth <- c(0,1,10,30,50)
alewife.spawning$depth.SIV <- c(0.5,1,1,0,0)
alewife.spawning$salinity <- c(0,5,10,25,35)
alewife.spawning$salinity.SIV <- c(1,1,0.5,0,0)
alewife.spawning$temp.C <- c(8,12,22,30,45)
alewife.spawning$temp.C.SIV <- c(0,1,1,0,0)
alewife.spawning$velocity <- c(0,0.05,0.3,1,5)
alewife.spawning$velocity.SIV <- c(0,1,1,0,0)
alewife.spawning$silt.clay.per <- c(0,0.2,0.4,0.8,1)
alewife.spawning$silt.clay.SIV <- c(1,1,0.5,0,0)
alewife.spawning$tubidity <- c(0,0.2,0.4,0.8,1)
alewife.spawning$turbidity.SIV <- c(1,1,1,0.5,0)

##########
#Create summary figure for fresh.tid submodel
par(mfrow=c(2,3), cex=1.0)

  #Depth
  plot(alewife.spawning$depth, alewife.spawning$depth.SIV, xlim=c(0,100), ylim=c(0,1), type="l", xlab="Depth (meters)", ylab="Suitability Index (depth)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.spawning$depth, alewife.spawning$depth.SIV, pch=19, cex=2)
  lines(alewife.spawning$depth, alewife.spawning$depth.SIV, lwd=3)
  
  #salinity.
  plot(alewife.spawning$salinity, alewife.spawning$salinity.SIV, xlim=c(0,35), ylim=c(0,1), type="l",
       xlab="Salinity (psu)", ylab="Suitability Index (salinity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.spawning$salinity, alewife.spawning$salinity.SIV, pch=19, cex=2)
  lines(alewife.spawning$salinity, alewife.spawning$salinity.SIV, lwd=3)
  
  #temperature
  plot(alewife.spawning$temp.C, alewife.spawning$temp.C.SIV, xlim=c(0,50), ylim=c(0,1), type="l",
       xlab="Temperature (Celcius)", ylab="Suitability Index (temperature)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.spawning$temp.C, alewife.spawning$temp.C.SIV, pch=19, cex=2)
  lines(alewife.spawning$temp.C, alewife.spawning$temp.C.SIV, lwd=3)
  
  #velocity
  plot(alewife.spawning$velocity, alewife.spawning$velocity.SIV, xlim=c(0,8), ylim=c(0,1), type="l",
       xlab="Velocity (m/s)", ylab="Suitability Index (velocity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.spawning$velocity, alewife.spawning$velocity.SIV, pch=19, cex=2)
  lines(alewife.spawning$velocity, alewife.spawning$velocity.SIV, lwd=3)
  
  #silt/clay percentage
  plot(alewife.spawning$silt.clay.per, alewife.spawning$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.spawning$silt.clay.per, alewife.spawning$silt.clay.SIV, pch=19, cex=2)
  lines(alewife.spawning$silt.clay.per, alewife.spawning$silt.clay.SIV, lwd=3)
  
   #turbidity
  plot(alewife.spawning$silt.clay.per, alewife.spawning$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.spawning$silt.clay.per, alewife.spawning$silt.clay.SIV, pch=19, cex=2)
  lines(alewife.spawning$silt.clay.per, alewife.spawning$silt.clay.SIV, lwd=3)
```

#### Depth

According to Brown et al (2000), spawning adult alewives have a depth preference for MLT-10 meters. There is also a lower preference for habitat greater than 30 meters in depth (Pardue). In (Mather et al., 2012), 75% of alewives were observed in less than 2 meters. According to (O’Connell & Angermeier, 1997), Alewives spawn in both shallow and deep water.

#### Salinity
Brown et al (2000) also describes a higher tolerance for salinity with high preference for habitat with salinity concentrations of <15 psu. Salinty concentrations greater than 20 are unsuitable for spawning adults.

#### Temperature
Brown et al (2000) describes a wide range of temperature preferences with optimal temperatures at 12-16 degrees celcius. The total range of temperatures suitable for spawning alewives is 8-20 degrees celcuis. Spawning success did not occur past 29.7 degrees (Pardue). Alewife migrate into freshwater streams when water temperatures reach between 8-18 degrees celcius (Mather et al., 2012). Spawning ceases at water temperatures exceeding 27◦C (Kissil 1974). Appropriate spawning temperatures broadly fall between 10◦C and 22◦C (Tyus 1974; Pardue 1983;Collette  and  Klein-MacPhee  2002). Temperatures from 11-28 are suitable for alewife eggs (Klauda et al. 1991).

#### Flow Velocity
According to O'connell 1999, habitat patches with velocity values between .06 m/s and 0.16 m/s have a probability of 0.5 for the presence alewife eggs. Pardue describes velocities less than 0.3 m/s as suitable habitat for spawning.

#### Sediment Composition/ grain size distribution
Spawn over a range of substrates like gravel, sand, and other soft substrates (O’Connell & Angermeier, 1997).

#### Turbity 
Spawning is extremely tolerant of suspended sediments (Pardue);(Auld and Schubel 1978).

## Juvenile Habitat
```{r echo=FALSE, fig.width=14, fig.height=10, fig.cap="Suitability index curves for Juvenile Alewife Habitat"}
#Create data frame with freshwater tidal submodel
alewife.juv <- data.frame(matrix(NA,nrow=5,ncol=12))
colnames(alewife.juv) <- c("depth","depth.SIV","salinity", "salinity.SIV", "temp.C", "temp.C.SIV","velocity","velocity.SIV","silt.clay.per","silt.clay.SIV","tubidity","turbidity.SIV")

alewife.juv$depth <- c(0,1,10,20,50)
alewife.juv$depth.SIV <- c(0.5,1,1,0,0)
alewife.juv$salinity <- c(0,5,10,25,35)
alewife.juv$salinity.SIV <- c(0.1,0.5,1,1,0.5)
alewife.juv$temp.C <- c(6,12,22,24,30)
alewife.juv$temp.C.SIV <- c(0.1,1,1,0.5,0)
alewife.juv$velocity <- c(0,0.05,0.3,1,5)
alewife.juv$velocity.SIV <- c(0,1,1,0,0)
alewife.juv$silt.clay.per <- c(0,0.2,0.4,0.8,1)
alewife.juv$silt.clay.SIV <- c(1,1,0.5,0,0)
alewife.juv$tubidity <- c(0,0.2,0.4,0.8,1)
alewife.juv$turbidity.SIV <- c(1,1,0.5,0,0)

##########
#Create summary figure for fresh.tid submodel
par(mfrow=c(2,3), cex=1.0)

  #Depth
  plot(alewife.juv$depth, alewife.juv$depth.SIV, xlim=c(0,100), ylim=c(0,1), type="l", xlab="Depth (meters)", ylab="Suitability Index (depth)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.juv$depth, alewife.juv$depth.SIV, pch=19, cex=2)
  lines(alewife.juv$depth, alewife.juv$depth.SIV, lwd=3)
  
  #salinity.
  plot(alewife.juv$salinity, alewife.juv$salinity.SIV, xlim=c(0,35), ylim=c(0,1), type="l",
       xlab="Salinity (psu)", ylab="Suitability Index (salinity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.juv$salinity, alewife.juv$salinity.SIV, pch=19, cex=2)
  lines(alewife.juv$salinity, alewife.juv$salinity.SIV, lwd=3)
  
  #temperature
  plot(alewife.juv$temp.C, alewife.juv$temp.C.SIV, xlim=c(0,50), ylim=c(0,1), type="l",
       xlab="Temperature (Celcius)", ylab="Suitability Index (temperature)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.juv$temp.C, alewife.juv$temp.C.SIV, pch=19, cex=2)
  lines(alewife.juv$temp.C, alewife.juv$temp.C.SIV, lwd=3)
  
  #velocity
  plot(alewife.juv$velocity, alewife.juv$velocity.SIV, xlim=c(0,8), ylim=c(0,1), type="l",
       xlab="Velocity (m/s)", ylab="Suitability Index (velocity)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.juv$velocity, alewife.juv$velocity.SIV, pch=19, cex=2)
  lines(alewife.juv$velocity, alewife.juv$velocity.SIV, lwd=3)
  
  #silt/clay percentage
  plot(alewife.juv$silt.clay.per, alewife.juv$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.juv$silt.clay.per, alewife.juv$silt.clay.SIV, pch=19, cex=2)
  lines(alewife.juv$silt.clay.per, alewife.juv$silt.clay.SIV, lwd=3)
  
   #turbidity
  plot(alewife.juv$silt.clay.per, alewife.juv$silt.clay.SIV, xlim=c(0,1), ylim=c(0,1), type="l",
       xlab="Silt/Clay Percent", ylab="Suitability Index (sediment composition)")
  abline(h=seq(0,1,0.1), lty=3)
  points(alewife.juv$silt.clay.per, alewife.juv$silt.clay.SIV, pch=19, cex=2)
  lines(alewife.juv$silt.clay.per, alewife.juv$silt.clay.SIV, lwd=3)
```

#### Depth
Depth preference for juvenile alewives was very similar to spawning adults with a preference of 0-10 meters with no habitat suitability past 20 meters (Brown et al (2000)).

#### Salinity
Salinity preferences are higher for juvenile alewives than spawning adults. Juvenile alewives were listed to have a high preference for habitat with salinity concentrations >10 psu but show some suitability for freshwater tidal environments (Brown et al (2000)). Can even be found in concentraions up to 30 psu (Pardue).

#### Temperature
Temperature preferences for juvenile alewives is a wider range than spawning adults. Juvenile alewives have a preference for temperatures from 6-30 degrees celcius with optimal suitability from 12-22 degrees Celsius (Brown et al (2000)). Pardue also verifies preference for lower temperatures around 6 degrees Celsius. Temperature also verified in where 100% survival is shown for 15 psu and decreased survival is experiences around 0 psu and 30 psu (DiMaggio 2015).

#### Flow Velocity

#### Sediment Composition

#### Turbity 

## References